---
title: "The Fiscal Cost of the Sprawl Paradox"
subtitle: "Transportation Investment, Property Tax Productivity, and the Cross-Subsidy of Low-Density Land — OKI Region"
author: "Benedict Leonardi"
date: today
execute:
  echo: false
  warning: false
---

## Overview

---

## Data and Methodology

### Fiscal Productivity

$$\text{Revenue}_{i} \approx \text{Median Home Value}_i \times \text{Housing Units}_i \times 0.0125$$

### Federal Investment

Transportation investment is measured using federal award data from USASpending.gov, with awards geocoded to Census block groups and aggregated to tracts. We focus on DOT-sourced awards (CFDA prefixes 20.xxx) to isolate transportation-specific investment.

### Density Classification

Tracts are classified into three density tiers consistent with the TUC framework:

| Tier | Density Threshold | Description |
|------|-------------------|-------------|
| Urban | ≥ 5,000 ppsm | Continuous built fabric; transit-viable |
| Suburban | 1,000–5,000 ppsm | Mixed residential; auto-dependent |
| Sprawl | < 1,000 ppsm | Low-density; high infrastructure-per-capita cost |

---

## Fiscal Productivity by Density Class

```{r summary-table}
library(tidyverse)
library(scales)
library(knitr)

roi_dir <- "../Transportation_ROI_Analysis/data"
results <- read_csv(file.path(roi_dir, "oki_roi_results.csv"), show_col_types = FALSE) %>%
  mutate(GEOID = str_pad(as.character(GEOID), 12, pad = "0"))
summary <- read_csv(file.path(roi_dir, "oki_roi_summary.csv"), show_col_types = FALSE)

summary %>%
  kable(
    digits = 2,
    caption = "Fiscal and Investment Metrics by Density Tier (OKI Region, per acre)",
    col.names = c("Density Tier", "Tracts", "Avg. Revenue/Acre", "Avg. Fed. Investment/Acre", "Avg. Maintenance Liability/Acre", "Net Fiscal Impact/Acre")
  )
```

---

## Spatial Distribution of ROI

```{r map}
library(sf)
library(leaflet)

tracts <- st_read(file.path(roi_dir, "oki_tracts.geojson"), quiet = TRUE) %>%
  mutate(GEOID = str_pad(as.character(GEOID), 12, pad = "0"))

map_data <- tracts %>% inner_join(results, by = "GEOID")

pal <- colorNumeric(palette = "YlOrRd", domain = map_data$Revenue_Per_Acre)

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(Revenue_Per_Acre),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    label = ~paste0(NAME, ": ", dollar(Revenue_Per_Acre), "/acre")
  ) %>%
  addLegend(
    pal = pal, values = ~Revenue_Per_Acre,
    opacity = 0.7,
    title = "Property Tax Revenue / Acre",
    position = "bottomright"
  )
```

---

## The Cross-Subsidy Structure

---

## Methodological Notes

- Revenue estimates assume a uniform 1.25% effective tax rate. Actual rates vary substantially across Hamilton, Kenton, Campbell, and the five other OKI counties; the proxy is suitable for cross-sectional comparison within the region but not for absolute revenue projection.
- Federal award geocoding maps point awards to the nearest road segment using Overture Maps data, then aggregates to Census geographies. Awards with county-level precision only are excluded.
- Maintenance liability estimates follow AASHTO's standard $10,000/lane-mile/year figure, applied to total road mileage per block group from OpenStreetMap.

---

## References

- Federal Highway Administration (2024). *Highway Statistics Series.*
- USASpending.gov (2023). *Federal Award Data, DOT Transportation Programs.*
- U.S. Census Bureau (2022). *American Community Survey 5-Year Estimates.*
- Overture Maps Foundation (2025). *Roads Layer: Technical Documentation.*
- Strong Towns (2021). *The Cost of Car Dependency.*
