---
title: "The Fiscal Cost of the Sprawl Paradox"
subtitle: "Transportation Investment, Property Tax Productivity, and the Cross-Subsidy of Low-Density Land — OKI Region"
author: "Benedict Leonardi"
date: today
execute:
  echo: false
  warning: false
  eval: false
---

## Overview

If the True Urban Core framework exposes a measurement problem — official city populations overstating functional urban density — then this analysis exposes the downstream cost: federal transportation investment is systematically misallocated along the same fault line.

The OKI region (the 8-county Ohio-Kentucky-Indiana area surrounding Cincinnati) offers an unusually clean test case. The Ohio River bisects a continuous urban fabric along a state boundary, creating the same official-vs.-actual boundary distortion examined in the main paper. Within this region, we can match tract-level fiscal productivity against tract-level federal infrastructure investment and ask: are the places generating the most public revenue receiving their proportional share of public spending?

The answer is no — by a factor of roughly 17.

---

## Data and Methodology

### Fiscal Productivity

We estimate property tax productivity at the Census tract level using a revenue proxy:

$$\text{Revenue}_{i} \approx \text{Median Home Value}_i \times \text{Housing Units}_i \times 0.0125$$

This approximation applies a uniform 1.25% effective tax rate to the total estimated residential stock. It captures the relative fiscal productivity of tracts — the ratio of urban to sprawl productivity — rather than precise absolute figures.

### Federal Investment

Transportation investment is measured using federal award data from USASpending.gov, with awards geocoded to Census block groups and aggregated to tracts. We focus on DOT-sourced awards (CFDA prefixes 20.xxx) to isolate transportation-specific investment.

### Density Classification

Tracts are classified into three density tiers consistent with the TUC framework:

| Tier | Density Threshold | Description |
|------|-------------------|-------------|
| Urban | ≥ 5,000 ppsm | Continuous built fabric; transit-viable |
| Suburban | 1,000–5,000 ppsm | Mixed residential; auto-dependent |
| Sprawl | < 1,000 ppsm | Low-density; high infrastructure-per-capita cost |

---

## Fiscal Productivity by Density Class

```{r summary-table}
library(tidyverse)
library(scales)
library(knitr)

results <- read_csv("data/oki_roi_results.csv", show_col_types = FALSE) %>%
  mutate(GEOID = str_pad(as.character(GEOID), 12, pad = "0"))
summary <- read_csv("data/oki_roi_summary.csv", show_col_types = FALSE)

summary %>%
  kable(
    digits = 2,
    caption = "Fiscal and Investment Metrics by Density Tier (OKI Region, per acre)",
    col.names = c("Density Tier", "Tracts", "Avg. Revenue/Acre", "Avg. Fed. Investment/Acre", "Avg. Maintenance Liability/Acre")
  )
```

Urban tracts generate approximately **17 times** more estimated property tax revenue per acre than sprawl-class tracts. Yet the federal investment ranking runs in the opposite direction: suburban tracts receive the most investment per acre, while urban tracts receive the least.

This inversion is not coincidental. It reflects the structure of federal transportation programs, which historically fund lane-miles rather than people-per-acre. A highway interchange in a low-density suburb consumes more acreage — and therefore more federal dollars — than the same investment in urban infrastructure that serves a vastly larger population within the same footprint.

---

## Spatial Distribution of ROI

```{r map}
library(sf)
library(leaflet)

tracts <- st_read("data/oki_tracts.geojson", quiet = TRUE) %>%
  mutate(GEOID = str_pad(as.character(GEOID), 12, pad = "0"))

map_data <- tracts %>% inner_join(results, by = "GEOID")

pal <- colorNumeric(palette = "YlOrRd", domain = map_data$Revenue_Per_Acre)

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    fillColor = ~pal(Revenue_Per_Acre),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    label = ~paste0(NAME, ": ", dollar(Revenue_Per_Acre), "/acre")
  ) %>%
  addLegend(
    pal = pal, values = ~Revenue_Per_Acre,
    opacity = 0.7,
    title = "Property Tax Revenue / Acre",
    position = "bottomright"
  )
```

The spatial pattern is striking but unsurprising: the highest revenue-per-acre tracts cluster in the Cincinnati urban core, Covington, and Newport — precisely the area that the main paper identifies as functionally continuous urban fabric despite being split across official municipal and state boundaries. The lowest-productivity tracts radiate outward into the exurban counties that nonetheless represent large shares of the region's total road network.

---

## The Cross-Subsidy Structure

The combined picture — high urban fiscal productivity, low urban infrastructure investment — implies a structural cross-subsidy flowing from dense cores to low-density peripheries. This is not unique to Cincinnati. It is the standard operating condition of American metropolitan public finance:

1. Urban tracts generate the majority of the property tax base and, via agglomeration, drive the wage and sales tax base.
2. Infrastructure programs are structured around vehicle-miles and lane-miles, which scale with land area rather than population density.
3. Low-density development generates large maintenance liabilities — approximately **$10,000 per lane-mile annually** — that are partly offset by transfers from higher-productivity jurisdictions.

The implication for the True Urban Core framework is direct: cities that score high on official population rankings but low on TUC-adjusted urbanicity (Indianapolis, Columbus, Phoenix) are extracting cross-subsidies from the dense regions they nominally outrank. The "Sprawl Paradox" in population measurement has an exact fiscal analog in transportation spending.

---

## Methodological Notes

- Revenue estimates assume a uniform 1.25% effective tax rate. Actual rates vary substantially across Hamilton, Kenton, Campbell, and the five other OKI counties; the proxy is suitable for cross-sectional comparison within the region but not for absolute revenue projection.
- Federal award geocoding maps point awards to the nearest road segment using Overture Maps data, then aggregates to Census geographies. Awards with county-level precision only are excluded.
- Maintenance liability estimates follow AASHTO's standard $10,000/lane-mile/year figure, applied to total road mileage per block group from OpenStreetMap.

---

## References

- Federal Highway Administration (2024). *Highway Statistics Series.*
- USASpending.gov (2023). *Federal Award Data, DOT Transportation Programs.*
- U.S. Census Bureau (2022). *American Community Survey 5-Year Estimates.*
- Overture Maps Foundation (2025). *Roads Layer: Technical Documentation.*
- Strong Towns (2021). *The Cost of Car Dependency.*
