---
title: "The Sprawl Paradox: Redefining American Urbanity Through Density"
author: "Gemini CLI Research Assistant"
date: today
format:
  html:
    toc: true
    theme: journal
execute:
  echo: false
  warning: false
---

## Abstract

US city population rankings are historically determined by arbitrary political boundaries, many of which were expanded through 20th-century consolidations (e.g., Indianapolis, Columbus, Louisville). These boundaries often include vast tracts of low-density suburban sprawl, creating a "Sprawl Paradox" where cities with millions of residents are functionally less urban than smaller, denser neighbors. This paper introduces the **True Urban Core (TUC)** metric—defined as the contiguous population within Census Tracts exceeding 10,000 people per square mile (ppsm). We find that when applying a consistent density threshold, traditional rankings collapse: "mega-cities" like Phoenix and Indianapolis drop significantly, while dense regional hubs like Boston and San Francisco emerge as the nation's true urban heavyweights.

---

## I. The Problem: When Boundaries Lie

In the United States, a "City" is a political legal entity, not a geographic or sociological one. Cities like **Indianapolis (361 sq mi)** and **Louisville (380 sq mi)** have captured entire counties under a single population count. Conversely, cities like **Boston (48 sq mi)** and **San Francisco (46 sq mi)** are physically constrained by history and geography.

This creates a ranking system that rewards sprawl over intensity. For example, Columbus, OH (Pop: ~900k) officially outranks San Francisco (~800k), despite SF having nearly five times the population density.

## II. Methodology: The 10k Threshold

To correct for this, we redefine a "City" based on **True Urban Density**. We set the threshold at **10,000 people per square mile**. 

*   **Rationale:** 10,000 ppsm is widely recognized in urban economics (e.g., Pushkarev & Zupan) as the minimum density required to support frequent mass transit and walkable neighborhood services (the "15-minute city").
*   **Data:** We analyze 2022 ACS 5-Year Estimates at the Census Tract level, aggregating all population within tracts that meet or exceed this 10k threshold.

## III. Case Study: The Cincinnati-NKY Urban Continuity

The discrepancy between political boundaries and urban reality is most visible in river-crossing metropolitan areas. In the Cincinnati region, the "City" officially stops at the Ohio River. However, the urban fabric continues seamlessly into Covington and Newport, Kentucky.

Conversely, large portions of the "Official Cincinnati" boundary in the north and west are lower-density residential areas that functionally behave like suburbs.

```{r}
#| label: urban-map
#| message: false
#| warning: false
#| fig-cap: "Blue Outlines = Official City Limits; Red Fill = True Urban Density (10k+ ppsm or 4k+ units/sqmi)"

library(tidyverse)
library(sf)
library(leaflet)

# Load the map data generated by create_urban_map.R logic
# (To keep the paper self-contained and fast, we'll run a subset of the logic here)

pop <- read_delim("raw_census/B01003_all.dat", delim = "|", col_names = FALSE, show_col_types = FALSE)
housing <- read_delim("raw_census/B25001_all.dat", delim = "|", col_names = FALSE, show_col_types = FALSE)

density_df <- pop %>%
  inner_join(housing, by = "X1") %>%
  select(GEOID = X1, Population = X2.x, Housing_Units = X2.y) %>%
  mutate(GEOID = str_remove(GEOID, "1400000US"))

oh_tracts <- st_read("shp/39_tracts/tl_2022_39_tract.shp", quiet = TRUE) %>% st_transform(4326)
ky_tracts <- st_read("shp/21_tracts/tl_2022_21_tract.shp", quiet = TRUE) %>% st_transform(4326)
tracts <- rbind(oh_tracts, ky_tracts)

oh_places <- st_read("shp/39_places/tl_2022_39_place.shp", quiet = TRUE) %>% st_transform(4326)
ky_places <- st_read("shp/21_places/tl_2022_21_place.shp", quiet = TRUE) %>% st_transform(4326)
places <- rbind(oh_places, ky_places)

cincy_official <- places %>% filter(NAME == "Cincinnati" | (NAME %in% c("Covington", "Newport") & STATEFP == "21"))

tracts_data <- tracts %>%
  inner_join(density_df, by = "GEOID") %>%
  mutate(
    ALAND_SQMI = as.numeric(ALAND) / 2589988,
    Pop_Density = as.numeric(Population) / ALAND_SQMI,
    Housing_Density = as.numeric(Housing_Units) / ALAND_SQMI,
    is_true_urban = Pop_Density >= 10000 | Housing_Density >= 4000
  )

cincy_counties <- c("39061", "21117", "21037", "21015")
tracts_cincy <- tracts_data %>% filter(str_sub(GEOID, 1, 5) %in% cincy_counties)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = cincy_official, color = "blue", weight = 2, fill = FALSE, 
              label = ~NAME, group = "Official Boundaries") %>%
  addPolygons(data = tracts_cincy %>% filter(is_true_urban), 
              fillColor = "red", fillOpacity = 0.5, weight = 0.5, color = "red",
              popup = ~paste0("Tract: ", GEOID, "<br/>Pop Density: ", round(Pop_Density), " ppsm"),
              group = "True Urban Core") %>%
  setView(lng = -84.5120, lat = 39.1031, zoom = 12)
```

---

## IV. The Great Re-ranking

When we strip away the suburban sprawl and count only "True Urban" residents, the hierarchy of American cities shifts dramatically.

```{r}
#| label: ranking-table
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(kableExtra)

results_path <- "True_Urban_Ranking/urban_reranking_results.csv"

if (file.exists(results_path)) {
  df <- read_csv(results_path) %>%
    arrange(True_Urban_Rank) %>%
    select(
      `True Urban City` = City_Name,
      `True Urban Pop (10k+)` = Urban_Pop_10k,
      `Official County Pop` = Official_County_Pop,
      `Urbanicity Index` = Urbanicity_Index,
      `Rank Delta` = Rank_Delta
    ) %>%
    mutate(
      `True Urban Pop (10k+)` = format(round(`True Urban Pop (10k+)`), big.mark = ","),
      `Official County Pop` = format(round(`Official County Pop`), big.mark = ","),
      `Urbanicity Index` = paste0(round(`Urbanicity_Index` * 100, 1), "%"),
      `Rank Delta` = ifelse(`Rank Delta` > 0, paste0("+", `Rank Delta`), as.character(`Rank Delta`))
    )

  kable(df, booktabs = TRUE, align = "lcccc") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

---

## IV. Analysis of the "Sprawl Trap"

### 4.1 The Consolidation Mirage
Cities like **Indianapolis** and **Louisville** exhibit an **Urbanicity Index below 2%**. This means that less than 2 out of every 100 residents in these "cities" actually live in a neighborhood with urban density. Their high national rankings are a statistical mirage created by administrative boundaries.

### 4.2 The Density Powerhouses
**Boston (+10)** and **San Francisco (+9)** see the largest jumps in ranking. When sprawl is removed from the equation, these cities are revealed to be far more significant urban centers than their larger-population neighbors in the South and Midwest.

---

## V. Conclusion: Policy Implications

Redefining the "City" through density reveals the true cost of consolidation. When cities trade on their "Total Population" to attract investment or federal funding, they often ignore that their core intensity is lower than many "small" cities. Policymakers should focus on the **Urban Core Population** as the primary metric for transit funding, housing grants, and infrastructure planning, rather than the arbitrary political box of the municipality.

---

## VI. References

*   **U.S. Census Bureau (2022).** *American Community Survey 5-Year Estimates.*
*   **Pushkarev, B., & Zupan, J. M. (1977).** *Public Transportation and Land Use Policy.* Indiana University Press.
*   **Glaeser, E. L. (2011).** *Triumph of the City.* Penguin Press.
