---
title: "The Sprawl Paradox: Redefining American Urbanity Through Density"
author: "Benedict Leonardi"
date: today
execute:
  echo: false
  warning: false
---

## Abstract

---

## I. The Problem: When Boundaries Lie

---

## II. Methodology: The 10,000 ppsm Threshold

The [5K Analysis](URBAN_RERANKING_PAPER_5K.html) page applies the same methodology at the 5,000 ppsm threshold — the Census Bureau's standard definition of "urbanized" — for comparison.

---

## III. Case Study: The Cincinnati–NKY Urban Continuity

```{r}
#| label: urban-map
#| message: false
#| warning: false
#| fig-cap: "Blue outlines = official city limits. Red fill = True Urban Core (≥10k ppsm or ≥4k units/sq mi)."

library(tidyverse)
library(sf)
library(leaflet)

pop <- read_delim("raw_census/B01003_all.dat", delim = "|", col_names = FALSE, show_col_types = FALSE)
housing <- read_delim("raw_census/B25001_all.dat", delim = "|", col_names = FALSE, show_col_types = FALSE)

density_df <- pop %>%
  inner_join(housing, by = "X1") %>%
  select(GEOID = X1, Population = X2.x, Housing_Units = X2.y) %>%
  mutate(GEOID = str_remove(GEOID, "1400000US"))

oh_tracts <- st_read("shp/39_tracts/tl_2022_39_tract.shp", quiet = TRUE) %>% st_transform(4326)
ky_tracts <- st_read("shp/21_tracts/tl_2022_21_tract.shp", quiet = TRUE) %>% st_transform(4326)
tracts <- rbind(oh_tracts, ky_tracts)

oh_places <- st_read("shp/39_places/tl_2022_39_place.shp", quiet = TRUE) %>% st_transform(4326)
ky_places <- st_read("shp/21_places/tl_2022_21_place.shp", quiet = TRUE) %>% st_transform(4326)
places <- rbind(oh_places, ky_places)

cincy_official <- places %>% filter(NAME == "Cincinnati" | (NAME %in% c("Covington", "Newport") & STATEFP == "21"))

tracts_data <- tracts %>%
  inner_join(density_df, by = "GEOID") %>%
  mutate(
    ALAND_SQMI = as.numeric(ALAND) / 2589988,
    Pop_Density = as.numeric(Population) / ALAND_SQMI,
    Housing_Density = as.numeric(Housing_Units) / ALAND_SQMI,
    is_true_urban = Pop_Density >= 10000 | Housing_Density >= 4000
  )

cincy_counties <- c("39061", "21117", "21037", "21015")
tracts_cincy <- tracts_data %>% filter(str_sub(GEOID, 1, 5) %in% cincy_counties)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = cincy_official, color = "blue", weight = 2, fill = FALSE,
              label = ~NAME, group = "Official Boundaries") %>%
  addPolygons(data = tracts_cincy %>% filter(is_true_urban),
              fillColor = "red", fillOpacity = 0.5, weight = 0.5, color = "red",
              popup = ~paste0("Tract: ", GEOID, "<br/>Pop Density: ", round(Pop_Density), " ppsm"),
              group = "True Urban Core") %>%
  setView(lng = -84.5120, lat = 39.1031, zoom = 12)
```

See also the [Transit Fragmentation](https://bleonardi.github.io/Transit_Fragmentation_Research/) research.

---

## IV. The Great Re-ranking

```{r}
#| label: ranking-table
#| message: false
#| warning: false

library(knitr)
library(kableExtra)

results_path <- "True_Urban_Ranking/urban_reranking_results.csv"

if (file.exists(results_path)) {
  df <- read_csv(results_path) %>%
    arrange(True_Urban_Rank) %>%
    select(
      `True Urban City` = City_Name,
      `True Urban Pop (10k+)` = Urban_Pop_10k,
      `Official County Pop` = Official_County_Pop,
      `Urbanicity Index` = Urbanicity_Index,
      `Rank Delta` = Rank_Delta
    ) %>%
    mutate(
      `True Urban Pop (10k+)` = format(round(`True Urban Pop (10k+)`), big.mark = ","),
      `Official County Pop` = format(round(`Official County Pop`), big.mark = ","),
      `Urbanicity Index` = paste0(round(`Urbanicity_Index` * 100, 1), "%"),
      `Rank Delta` = ifelse(`Rank Delta` > 0, paste0("+", `Rank Delta`), as.character(`Rank Delta`))
    )

  kable(df, booktabs = TRUE, align = "lcccc") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

---

## V. Interpreting the Results

### The Consolidation Mirage

### The Density Powerhouses

### The Fiscal Implication

See the [Fiscal ROI analysis](fiscal_roi.html).

---

## VI. Conclusion

---

## References

- U.S. Census Bureau (2022). *American Community Survey 5-Year Estimates.*
- Pushkarev, B., & Zupan, J. M. (1977). *Public Transportation and Land Use Policy.* Indiana University Press.
- Glaeser, E. L. (2011). *Triumph of the City.* Penguin Press.
- Albouy, D., & Ehrlich, G. (2018). Housing productivity and the social cost of land-use restrictions. *Journal of Urban Economics,* 107, 101–120.


---

*[View source on GitHub](https://github.com/bleonardi/True_Urban_Ranking_Project){target="_blank"}*
